Semantic integration of physiology phenotypes: an application to the cellular phenotype ontology

### 2011-12-01 trunk/paper/drinking.tex

\begin{abstract}
  The systematic observation of phenotypes has become a crucial tool
  of functional genomics, and several large international projects are
  currently underway to identify and characterize the phenotypes that
  are associated with genotypes in several species. To integrate
  phenotype descriptions within and across species, phenotype
  ontologies have been developed. Applying ontologies to unify
  phenotype descriptions in the domain of physiology has been a
  particular challenge due to the high complexity of the underlying
  domain.
  Here, we present the outline of a theory for an ontology of
  physiology-related phenotypes. We provide a formal description of
  process attributes and relate them to the attributes of their parts
  and participants. We apply our theory to create the Cellular
  Phenotype Ontology (CPO). The CPO is an ontology of morphological
  and physiological phenotypic abnormalities of cells, cell components
  and cellular processes. Its prime application is the unification of
  cellular phenotype descriptions across species by providing terms
  and uniform definition patterns. The CPO can further be used for the
  annotation of observed abnormalities in domains, such as systems
  microscopy, in which cellular abnormalities are observed and for
  which no phenotype ontology has been created.  The CPO and
  associated source code is available on
  \url{http://cell-phenotype.googlecode.com}.
\end{abstract}

@@

\section{Acknowledgements}
Funding for RH is provided by the European Commission's 7th Framework
Programme, RICORDO project, grant number 248502. MAH is supported by
Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
Medical Informatics, Statistics and Epidemiology, University of
Leipzig.  Funding for GR is provided by the European Union's Seventh
Framework Programme (FP7/2007-2013) under grant agreement n°258068;
EU-FP7-Systems Microscopy NoE. GVG is funded by NIH grant number R01
HG004838-02.

>>>>

\begin{abstract}
  The systematic observation of phenotypes has become a crucial tool
  of functional genomics, and several large international projects are
  currently underway to identify and characterize the phenotypes that
  are associated with genotypes in several species. To integrate
  phenotype descriptions within and across species, phenotype
  ontologies have been developed. Applying ontologies to unify
  phenotype descriptions in the domain of physiology has been a
  particular challenge due to the high complexity of the underlying
  domain.  Here, we present the outline of a theory for an ontology of
  physiology-related phenotypes. We provide a formal description of
  process attributes and relate them to the attributes of their parts
  and participants. We apply our theory to create the Cellular
  Phenotype Ontology (CPO). The CPO is an ontology of morphological
  and physiological phenotypic abnormalities of cells, cell components
  and cellular processes. Its prime application is the unification of
  cellular phenotype descriptions across species by providing terms
  and uniform definition patterns. The CPO can further be used for the
  annotation of observed abnormalities in domains, such as systems
  microscopy, in which cellular abnormalities are observed and for
  which no phenotype ontology has been created.  The CPO and the
  source code we generated to create the CPO are available on
  \url{http://cell-phenotype.googlecode.com}.
\end{abstract}

@@

\section{Acknowledgements}
Funding for RH is provided by the European Commission's 7th Framework
Programme, RICORDO project, grant number 248502. MAH is supported by
Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
Medical Informatics, Statistics and Epidemiology, University of
Leipzig.  Funding for GR is provided by the European Union's Seventh
Framework Programme (FP7/2007-2013) under grant agreement number
258068, EU-FP7-Systems Microscopy NoE. GVG is funded by NIH grant
number R01 HG004838-02.
